Racial and Ethnic Differences in ETI: Outcomes

### 2025-07-22 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
logistic models for number of exacerbations

@@

>>>>

```{r}
# Need to do some re-cleaning for the dataset that includes exacerbations
# Pregnancies
preg <- which(encounter_w_hosps$pregnant == 1)
encounter_w_hosps <- encounter_w_hosps[-preg, ]
# Days from ETI start
early <- which(encounter_w_hosps$Days < -365.25 * 3)
late <- which(encounter_w_hosps$Days > 365.25 * 2)
encounter_w_hosps <- encounter_w_hosps[-early, ]
encounter_w_hosps <- encounter_w_hosps[-late, ]
# Lung function cutoffs
lower <- 10
upper <- 150
# Remove impossible lung function values
low_fev1 <- which(encounter_w_hosps$gli_fev1_ppred_rn < lower)
high_fev1 <- which(encounter_w_hosps$gli_fev1_ppred_rn > upper)
encounter_w_hosps$gli_fev1_ppred_rn[low_fev1] <- NA
encounter_w_hosps$gli_fev1_ppred_rn[high_fev1] <- NA
low_fvc <- which(encounter_w_hosps$gli_fvc_ppred_rn < lower)
high_fvc <- which(encounter_w_hosps$gli_fvc_ppred_rn > upper)
encounter_w_hosps$gli_fvc_ppred_rn[low_fvc] <- NA
encounter_w_hosps$gli_fvc_ppred_rn[high_fvc] <- NA
# Get some summary statistics by participant
t1_participant <- encounter_w_hosps %>%
  group_by(eDWID) %>%
  summarise(
    `Years With Data` = length(unique(reviewyear)),
    `Number of Encounters` = n(),
    `Number of Encounters Pre-ETI` = sum(Days <= 0),
    `Number of Encounters Post-ETI` = sum(Days > 0),
    Age = last(na.omit(encounterage[Days <= 0])),
    across(all_of(continuous_outcomes), \(x) last(na.omit(x[Days <= 0]))),
    Gender = names(sort(table(Gender), decreasing = TRUE))[1],
    Race = names(sort(table(Race), decreasing = TRUE))[1],
    Insurance = names(sort(table(Insurance), decreasing = TRUE))[1],
    .groups = "drop"
  )
no_post <-
  t1_participant$eDWID[t1_participant$`Number of Encounters Post-ETI` == 0]
no_post_visits <- which(encounter_w_hosps$eDWID %in% no_post)
encounter_w_hosps <- encounter_w_hosps[-no_post_visits, ]
# Pre vs. post variable
encounter_w_hosps$pre_post <- factor(encounter_w_hosps$Days > 0,
  levels = c(F, T), labels = c("Pre-ETI", "Post-ETI")
)
# Prior modulator use
prior_mods <- encounter_w_hosps %>%
  select(eDWID, Days, vx661comb, vx445comb, Vx770, Vx809comb) %>%
  filter(Days < 0) %>%
  group_by(eDWID) %>%
  summarise(prior_modulator = any(vx661comb == 1) | any(vx445comb == 1) |
    any(Vx770 == 1) | any(Vx809comb))
prior_mods$prior_modulator <-
  fct_na_value_to_level(factor(prior_mods$prior_modulator), level = "No")
prior_mods$prior_modulator <- factor(prior_mods$prior_modulator,
  levels = c(T, "No"),
  labels = c("Yes", "No")
)
encounter_w_hosps <-
  left_join(encounter_w_hosps, prior_mods, by = join_by(eDWID))
# On enzymes for pancreatic insufficiency
encounter_w_hosps$isOnEnzymes <-
  factor(encounter_w_hosps$isOnEnzymes, levels = 0:1, labels = c("No", "Yes"))
# Pseudomonas
encounter_w_hosps$pseudomonasaeruginosa[
  is.na(encounter_w_hosps$pseudomonasaeruginosa)
] <- 0
encounter_w_hosps$pseudomonasaeruginosa <-
  factor(encounter_w_hosps$pseudomonasaeruginosa,
    levels = 0:1, labels = c("No", "Yes")
  )
```

@@

```{r}
# Fit mixed model with change point
pex_mod <- glmer(
  PEx ~ Days + Race * pre_post + age_group +
    Insurance + prior_modulator + isOnEnzymes + pseudomonasaeruginosa +
    (1 | eDWID),
  family = "binomial",
  data = encounter_w_hosps
)
# Model plot
pop_dat <- data.frame(expand.grid(
  Years = -3:2, Race = unique(encounter$Race), age_group = "[18,30)",
  Insurance = "Private", prior_modulator = "Yes",
  isOnEnzymes = "Yes", pseudomonasaeruginosa = "No"
))
pop_dat$slope_change_years <- (pop_dat$Years > 0) * pop_dat$Years
pred <- predict(change_mod, re.form = NA, newdata = pop_dat, se.fit = T)
pop_dat$pred <- pred$fit
pop_dat$ll <- pop_dat$pred - (1.96 * pred$se.fit)
pop_dat$ul <- pop_dat$pred + (1.96 * pred$se.fit)
ggplot(pop_dat, aes(x = Years, y = pred, group = Race, colour = Race)) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  geom_line() +
  xlab("Years from ETI Start") +
  ylab("gli_fev1_ppred_rn") +
  theme_classic()
# Model summary
change_mod %>%
  gtsummary::tbl_regression(intercept = T) %>%
  as_gt() %>%
  fmt_number(decimals = 3)
# Table of predicted values
pop_dat %>%
  select(-slope_change_years) %>%
  rename(gli_fev1_ppred_rn = pred, ci.ll = ll, ci.ul = ul) %>%
  gt(caption = "Predicted Values at Various Timepoints") %>%
  fmt_number(decimals = 3)
# Contrasts
aa_pre <- c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0)
aa_change <- c(0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0)
hl_pre <- c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0)
hl_change <- c(0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0)
nhw_pre <- c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
nhw_change <- c(0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
other_pre <- c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0)
other_change <- c(0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1)
contr <- rbind(
  "Black/AA Pre-ETI" = aa_pre,
  "Black/AA Post-ETI" = aa_pre + aa_change,
  "H/L Pre-ETI" = hl_pre,
  "H/L Post-ETI" = hl_pre + hl_change,
  "NHW Pre-ETI" = nhw_pre,
  "NHW Post-ETI" = nhw_pre + nhw_change,
  "Other Pre-ETI" = other_pre,
  "Other Post-ETI" = other_pre + other_change
)
change_mod_slopes <- glht(change_mod, linfct = contr)
change_mod_slopes_ci <- confint(change_mod_slopes, calpha = univariate_calpha())
change_mod_slopes_ci$confint %>%
  as.data.frame() %>%
  gt(caption = "Slopes by Race", rownames_to_stub = T) %>%
  fmt_number(decimals = 3)
```
